Relational Parametricity for Algebraically Indexed Types

### 2012-05-29 paper/section-general-framework.tex
Add AUCTeX commands to deal with multifile document
Fill in some of the section on well-typed terms

The well-typed programs are defined in \autoref{fig:programs}. FIXME:
explain each of the rules.

\begin{figure*}[t]
  \centering
  \begin{mathpar}
    \inferrule* [right=Var]
    {\Delta \vdash \Gamma \isCtxt \\ x : A \in \Gamma}
    {\Delta; \Gamma \vdash x : A}

@@

    \inferrule* [right=Cond]
    {\Delta; \Gamma \vdash M : A + B \\
      \Delta; \Gamma, x : A \vdash N_1 : C \\
      \Delta; \Gamma, y : B \vdash N_2 : C}
    {\Delta; \Gamma \vdash \textrm{case}\ M\ \textrm{of}\ \textrm{inl}\ x.N_1; \textrm{inr}\ y.N_2 : C}

@@

For any typing context $\Delta \vdash \Gamma \isCtxt$, we define its
index-erasure semantics $\ctxtSem{\Gamma}$ by induction on its
structure: $\ctxtSem{\epsilon} = 1$ and $\ctxtSem{\Gamma, x : A} =
\ctxtSem{\Gamma} \times \tySem{A}$, where $1$ is the canonical
one-element set.

@@

For a well-typed program $\Delta; \Gamma \vdash M : A$, we define the
\emph{erasure interpretation} as a function $\tmSem{e} :
\ctxtSem{\Gamma} \to \tySem{A}$, completely ignoring the indexing
information. In light of \lemref{lem:tyeqsubst-erasure}, we define
this function directly on the syntax of well-typed terms, rather than
on the typing derivations. We define $\tmSem{e}$ by the following
clauses:
\begin{eqnarray*}
  \tmSem{x}\eta & = & \eta_x \\
  \tmSem{(M,N)}\eta & = & (\tmSem{M}\eta, \tmSem{N}\eta) \\
  \tmSem{\pi_1M}\eta & = & \pi_1(\tmSem{M}\eta) \\
  \tmSem{\pi_2M}\eta & = & \pi_2(\tmSem{M}\eta) \\
  \tmSem{\mathrm{inl}\ M}\eta & = & \mathrm{inl}\ (\tmSem{M}\eta) \\
  \tmSem{\mathrm{inr}\ M}\eta & = & \mathrm{inr}\ (\tmSem{M}\eta) \\
  \left\llbracket
    \begin{array}{l}
      \textrm{case}\ M\ \textrm{of}\\
      \textrm{inl}\ x.N_1;\\
      \textrm{inr}\ y.N_2
    \end{array}\right\rrbracket^{\mathit{tm}}\eta & = &
  \left\{
    \begin{array}{ll}
      \tmSem{N_1}(\eta,a) & \textrm{if }\tmSem{M}\eta = \mathrm{inl}(a) \\
      \tmSem{N_2}(\eta,b) & \textrm{if }\tmSem{M}\eta = \mathrm{inr}(b)
    \end{array}
  \right. \\
  \tmSem{\lambda x. M}\eta & = & \lambda v. \tmSem{M}(\eta, v) \\
  \tmSem{M N}\eta & = & (\tmSem{M}\eta) (\tmSem{N}\eta) \\
  \tmSem{\Lambda i.\ M}\eta & = & \tmSem{M}\eta \\
  \tmSem{M[e]}\eta & = & \tmSem{M}\eta
\end{eqnarray*}
In the above, we have written $\eta_x$ to denote the appropriate
projection from $\eta$ to get the position corresponding to the
variable $x$ in the context.

@@

We now state and prove the abstraction theorem for well-typed programs
in our system. The abstraction theorem states that the type erasure
interpretation of a well-typed program is related to itself by the
relational interpretation of its type. From this result we will be
able to deduce interesting properties of programs adhering to the
algebraically-indexed typing discipline in
\autoref{sec:instantiations}. % FIXME: spruce this up when the instantiations section is actually written.

To state and prove the abstraction theorem for open terms, we first
need to extend the relational interpretation of types to typing
contexts. The relational interpretation of contexts is defined by the
following clauses, where we have used the pairing operation
$-\relTimes-$ defined in \autoref{sec:relational-semantics}.
\begin{eqnarray*}
  \rsem{\epsilon}{\relEnv{E}}{\rho} & = & \{(*,*)\} \\
  \rsem{\Gamma, x : A}{\relEnv{E}}{\rho} & = & \rsem{\Gamma}{\relEnv{E}}\rho \relTimes \rsem{A}{\relEnv{E}}\rho
\end{eqnarray*}
The relational interpretation of contexts inherits from the relational
interpretation of types the property of interpretating the application
of simultaneous substitutions as composition:
\begin{lemma}\label{lem:ctxtsubst-rel}
  If $\Delta \vdash \Gamma \isCtxt$ and $\sigma : \Delta' \Rightarrow
  \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
  $\rsem{\sigma^*\Gamma}{\relEnv{E}}\rho =
  \rsem{\Gamma}{\relEnv{E}}{(\rho \circ \sigma^*)}$.
\end{lemma}

\begin{theorem}[Abstraction]
  If $\Delta; \Gamma \vdash M : A$, then for all $\rho \in
  \relEnv{E}(\Delta)$ and $\eta_1, \eta_2 \in \ctxtSem{\Gamma}$ such
  that $(\eta_1, \eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$, we have
  $(\tmSem{M}\eta_1, \tmSem{M}\eta_2) \in \rsem{A}{\relEnv{E}}\rho$.
\end{theorem}

@@

>>>>

We now present the rules for well-typed programs over the collection
of types we generated in
\autoref{sec:algebraically-indexed-types}. Each well-type program is
assigned an index-erasure semantics, building on the index-erasure
semantics of types we defined in
\autoref{sec:index-erasure-semantics}. Our main result
(\thmref{thm:abstraction}) is that the index-erasure semantics is
every well-typed program is related to itself in the relational
interpretation of its type: this is the abstraction theorem for every
calculus in our general framework. In the next section we will use
this theorem to derive several interesting results about the programs
expressible in instances of our general framework.

Well-typed programs are defined with respect to well-indexed typing
contexts, which are in turn defined with respect to an index
context. Well-indexed typing contexts with respect to an index context
$\Delta$ are sequences of variable/type pairs with no repeated
variable names such that each type is well-indexed with respect to
$\Delta$. Formally, well-indexed typing contexts are defined by the
following two rules:
\begin{mathpar}
  \inferrule*
  { }
  {\Delta \vdash \epsilon \isCtxt}

  \inferrule*
  {\Delta \vdash \Gamma \isCtxt \\ \Delta \vdash A \isType \\ x \not\in \Gamma}
  {\Delta \vdash \Gamma, x : A \isCtxt}
\end{mathpar}
Substitution by a simultaneous substitution extends to typing contexts
by performing substitution pointwise on each individual type.

The typing rules for our system define the well-typed terms with
respect to an index context $\Delta$ and a type context $\Delta \vdash
\Gamma \isCtxt$. The judgement $\Delta; \Gamma \vdash M : A$ is
defined in \autoref{fig:programs}. The equational theory on types is
incorporated into the type system via the rule \TirName{TyEq}, which
allows for a term that is judged to have type $A$ to also have any
other equal type $B$.
\begin{figure*}[t]
  \centering
  \begin{mathpar}
    \inferrule* [right=Var]
    {\Delta \vdash \Gamma \isCtxt \\ x : A \in \Gamma}
    {\Delta; \Gamma \vdash x : A}

@@

    \inferrule* [right=Case]
    {\Delta; \Gamma \vdash M : A + B \\
      \Delta; \Gamma, x : A \vdash N_1 : C \\
      \Delta; \Gamma, y : B \vdash N_2 : C}
    {\Delta; \Gamma \vdash \textrm{case}\ M\ \textrm{of}\ \textrm{inl}\ x.N_1; \textrm{inr}\ y.N_2 : C}

@@

@@

We assign an index-erasure semantics to any well-indexed typing
context $\Delta \vdash \Gamma \isCtxt$ by induction on its structure:
$\ctxtSem{\epsilon} = \{*\}$ and $\ctxtSem{\Gamma, x : A} =
\ctxtSem{\Gamma} \times \tySem{A}$. For a well-typed program $\Delta;
\Gamma \vdash M : A$, we define the \emph{erasure interpretation} as a
function $\tmSem{e} : \ctxtSem{\Gamma} \to \tySem{A}$, completely
ignoring the indexing information. In light of
\lemref{lem:tyeqsubst-erasure}, we define this function directly on
the syntax of well-typed terms, rather than on the typing
derivations. We define $\tmSem{e}$ by the following clauses:
\begin{displaymath}
  \begin{array}{@{\hspace{0em}}l@{\hspace{0.5em}}c@{\hspace{0.5em}}l}
    \tmSem{x}\eta & \isDefinedAs & \eta_x \\
    \tmSem{(M,N)}\eta & \isDefinedAs & (\tmSem{M}\eta, \tmSem{N}\eta) \\
    \tmSem{\pi_1M}\eta & \isDefinedAs & \pi_1(\tmSem{M}\eta) \\
    \tmSem{\pi_2M}\eta & \isDefinedAs & \pi_2(\tmSem{M}\eta) \\
    \tmSem{\mathrm{inl}\ M}\eta & \isDefinedAs & \mathrm{inl}\ (\tmSem{M}\eta) \\
    \tmSem{\mathrm{inr}\ M}\eta & \isDefinedAs & \mathrm{inr}\ (\tmSem{M}\eta) \\
    \left\llbracket
      \begin{array}{l}
        \textrm{case}\ M\ \textrm{of}\\
        \textrm{inl}\ x.N_1;\\
        \textrm{inr}\ y.N_2
      \end{array}\right\rrbracket^{\mathit{tm}}\eta & \isDefinedAs &
    \left\{
      \begin{array}{ll}
        \tmSem{N_1}(\eta,a) & \textrm{if }\tmSem{M}\eta = \mathrm{inl}(a) \\
        \tmSem{N_2}(\eta,b) & \textrm{if }\tmSem{M}\eta = \mathrm{inr}(b)
      \end{array}
    \right. \\
    \tmSem{\lambda x. M}\eta & \isDefinedAs & \lambda v. \tmSem{M}(\eta, v) \\
    \tmSem{M N}\eta & \isDefinedAs & (\tmSem{M}\eta) (\tmSem{N}\eta) \\
    \tmSem{\Lambda i.\ M}\eta & \isDefinedAs & \tmSem{M}\eta \\
    \tmSem{M[e]}\eta & \isDefinedAs & \tmSem{M}\eta    
  \end{array}
\end{displaymath}
In the above, we have written $\eta_x$ to denote the appropriate
projection from $\eta$ to get the value in the environment
corresponding to the variable $x$ in the context.

@@

We now state and prove the abstraction theorem for well-typed programs
in our system. To state and prove this theorem for open terms, we
first need to extend the relational interpretation of types to typing
contexts. The relational interpretation of contexts is defined by the
following clauses, where we have used the pairing operation
$-\relTimes-$ defined in \autoref{sec:relational-semantics}.
\begin{eqnarray*}
  \rsem{\epsilon}{\relEnv{E}}{\rho} & \isDefinedAs & \{(*,*)\} \\
  \rsem{\Gamma, x : A}{\relEnv{E}}{\rho} & \isDefinedAs & \rsem{\Gamma}{\relEnv{E}}\rho \relTimes \rsem{A}{\relEnv{E}}\rho
\end{eqnarray*}
The relational interpretation of contexts inherits from the relational
interpretation of types the property of interpretating the application
of simultaneous substitutions as composition:
\begin{lemma}\label{lem:ctxtsubst-rel}
  If $\Delta \vdash \Gamma \isCtxt$ and $\sigma : \Delta' \Rightarrow
  \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
  $\rsem{\sigma^*\Gamma}{\relEnv{E}}\rho =
  \rsem{\Gamma}{\relEnv{E}}{(\rho \circ \sigma^*)}$.
\end{lemma}

\begin{theorem}[Abstraction]\label{thm:abstraction}
  If $\Delta; \Gamma \vdash M : A$, then for all $\rho \in
  \relEnv{E}(\Delta)$ and $\eta_1, \eta_2 \in \ctxtSem{\Gamma}$ such
  that $(\eta_1, \eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$, we have
  $(\tmSem{M}\eta_1, \tmSem{M}\eta_2) \in \rsem{A}{\relEnv{E}}\rho$.
\end{theorem}

@@

%%% Local Variables:
%%% TeX-master: "paper"
%%% End:
